---
title: "Climate Change Module"
author: 'Finn Travers'
format:
  html:
    embed-resources: true
---

```{r message=FALSE}
library(tidyverse)
```

Question 7 - lots of NA in database

Question 9 - Correct interpretation of no increase since 1998? Does Q7 change shape of plot?

Question 10 - What is temporal resolution for dataset?

Question 10 - Properly splicing data together and plotting

Uploading to github

## Warm-up: Examining CO2 trends in R

-   Example from <http://climate.nasa.gov/vital-signs/carbon-dioxide/>
-   Raw data from <https://gml.noaa.gov/webdata/ccgg/trends/co2/co2_mm_mlo.txt>

In 1958, Dr. Charles David Keeling (1928-2005), a scientist at Scripps Institute of Oceanography, began collecting data on atmospheric CO2 concentration at the Mauna Loa Observatory located in Hawaii. This dataset allowed us to understand the degree to which climate change is human-caused through our burning of fossil fuels and release of CO2 into the atmosphere. Due to his scientific achievements, Dr. Keeling was awarded the National Medal of Science by President George W. Bush in 2002. This is the highest award for lifetime scientific achievement that can be granted in the U.S. Today, you get to analyze this same dataset, except that you have more data than was available to Dr. Keeling and his colleagues because your dataset extends up to the current time.

To read the code, you will use a new function called `read_table`. It is similar to `read_csv` except it looks for spaces between column entries rather than commas (remember that csv stands for comma-separated values). Others include `read_tsv`, which uses tabs to separate entries. You can discover the separation type by putting <https://gml.noaa.gov/webdata/ccgg/trends/co2/co2_mm_mlo.txt> into your web browser and examining the file. The code also includes an argument for comment, which denotes the character used to define a line in the file as being a comment rather than data or a header. The file also reveals that the column headers are on lines with a comment character, so they won't be read. You will use the argument `col_names` to define the headers manually. Finally, `NA` values are defined by the numbers -1 and -99, so they need to be defined using the `na` argument (otherwise, they will be read as numeric values).

```{r message=FALSE}

co2 <-  read_table("https://gml.noaa.gov/webdata/ccgg/trends/co2/co2_mm_mlo.txt", 
                  comment="#",
                  col_names = c("year", "month", "decimal_date",
                                "monthly_average",
                                "deseasonalized", "days", "sd_days",
                                "unc_month_mean"),
                  na = c("-1", "-99.99"))
co2
```

```{r}
ggplot(co2, aes(x = decimal_date, y = monthly_average)) + 
  geom_line() + 
  geom_line(aes(y = deseasonalized), color = "blue") +
  labs(x = "Year", y = "CO2 concentration (ppm)")
```

**Question 1:**

Describe the overall trend in the CO2 data.

**Answer 1:**

The trend in CO2 data has a increasing positive linear trend. This shows that as the year increases, CO2 concentration increases as well.

**Question 2:**

How does CO2 vary within a year? What month is it at max? Min? What might explain this sub-annual pattern? (you will need to write code and make a plot to answer this question)

```{r}
co2_month_avg <- co2 |> 
  group_by(month) |> 
  summarize(co2_month_avg = mean(monthly_average, na.rm = TRUE))

co2_month_avg
```

**Answer 2:**

The CO2 varies by increasing during the spring months and decreasing during the fall months. This is supported by CO2 being highest during May with a **362.39 ppm** average and lowest during September with a **356.75 ppm** average. This can be explained by the CO2 average dropping after May during the summer months, when plants photosynthesize the most and draw the most CO2 from the atmosphere, which also explains why CO2 is the lowest in September, at the end of the photosynthetic season right before plants begin to lose their leaves and release CO2 back into the atmosphere.

## Global Temperature Data

Current climate change affects many aspects of the environment, with socio-economic consequences. For example, a warmer climate can allow new diseases to be introduced and persist (e.g., West Nile became established in the United States after an unusually warm winter, which allowed the mosquitoes that carry the virus to survive and spread). We are concerned not only with the actual temperature but also with the rate at which it changes. Rapid changes make it more likely that species cannot adapt and will go extinct.

Each of the most recent years has been the warmest on record. In this section, we will analyze global mean temperature data.

Data from: <https://climate.nasa.gov/vital-signs/global-temperature/>

**Question 3:**

Describe the data set to the best of your ability, given the documentation provided.

-   Where is the data from?
-   Describe the class of each column and what units it is measured in.
-   What is the difference between "global temperature" and "global temperature anomaly"?

**Answer 3:**

This data is from NASA's Goddard Institute for Space Studies (GISS) and measures the Land-Ocean Temperature Index. The columns in the data set are year (YYYY), annual mean temperature anomaly (C), and lowess smoothing temperature anomaly (C). The difference between global temperature and global temperature anomaly is that the first measures the flat mean temperature while the second compares the temperature against a baseline temperature to better track the change over time from a certain point. Using anomaly is better for tracking climate change practices as it better shows change over time.

**Question 4:**

Construct the necessary R code to import and prepare for plotting the following data set: <https://data.giss.nasa.gov/gistemp/graphs/graph_data/Global_Mean_Estimates_based_on_Land_and_Ocean_Data/graph.txt>

You'll need to determine the file's delimiter and any comments or skips. You will also need to be sure that you have column names. You will not directly use the code above to read the CO2 data, but that code provides helpful tips for reading the data.

```{r}
LOTI <-  read_table("https://data.giss.nasa.gov/gistemp/graphs/graph_data/Global_Mean_Estimates_based_on_Land_and_Ocean_Data/graph.txt", 
                  comment="#",
                  skip = 5,
                  col_names = c("year", "loti_mean", "loti_smoothing"))
```

**Answer 4:**

```{r}
LOTI
```

**Question 5:**

Plot the trend in global mean temperatures over time. Describe what you see in the plot and how you interpret the patterns you observe.

```{r}
ggplot(LOTI, aes(x = year, y = loti_mean)) +
  geom_line()+
  geom_line(aes(y = loti_smoothing), color = "blue") +
  labs(x = "Year", y = "Global Mean Temperatue Anomaly")
```

**Answer 5:**

The graph shows that there is a significantly positive increase in global mean temperatures starting in the early 20th century. The temperature anomaly looks like it fluctuates every year or so but maintains a steady increase. The flow with smoothing cuts down on these fluctuations and creates a more readable and representative line of temperature anomaly.

## Evaluating the evidence for a "Pause" in warming?

The [2013 IPCC Report](https://www.ipcc.ch/pdf/assessment-report/ar5/wg1/WG1AR5_SummaryVolume_FINAL.pdf) included a tentative observation of a "much smaller increasing trend" in global mean temperatures since 1998 than was observed previously. This led to much discussion in the media about the existence of a "Pause" or "Hiatus" in global warming rates, as well as much research looking into where the extra heat could have gone. (Examples discussing this question include articles in [The Guardian](http://www.theguardian.com/environment/2015/jun/04/global-warming-hasnt-paused-study-finds), [BBC News](http://www.bbc.com/news/science-environment-28870988), and [Wikipedia](https://en.wikipedia.org/wiki/Global_warming_hiatus)).

You will use rolling averages to help you explore the evidence for a pause. Since you have not been provided instructions for calculating rolling means, the learning objective of this question is to practice finding the solution.

**Question 6:**

Use a search engine (e.g., Google) or a chat LLM to find out how to calculate a rolling average in R. What search term or chat LLM prompt did you use? What website or answer from the chat LLM did you end up using? How much did you need to modify the code from the website or chat LLM to answer the question successfully?

**Answer 6:**

I used Microsoft Copilot using the prompt "How to calculate rolling average in R" I modified the code by inputting my own data into the template.

```{r}
library(zoo)
rolling_avg <- rollmean(LOTI, k = 3, fill = NA)
rolling_avg
```

**Question 7:**

-   What is the meaning of "5-year average" vs. "annual average"?
-   Create a data frame from the annual temperature anomaly data (from Question 4) with three new columns: 5-year running averages, 10-year running averages, and 20-year running averages.

**Answer 7:**

-   The "5 year average" is mean of data over a 5 year period divided by 5 to get the average of that timespan. The annual average is simply the mean of the data over a single year (annually)

    ```{r}
    LOTI_rolling_avg <- LOTI |> 
      mutate(five_year_avg = rollmean(loti_mean, k = 5, fill = NA),
             ten_year_avg = rollmean(loti_mean, k = 10, fill = NA),
             twenty_year_avg = rollmean(loti_mean, k = 20, fill = NA))
    ```

**Question 8:**

Plot the different averages on the *same plot* and describe what differences you see and why.

**Answer 8:**

```{r}
ggplot(data = LOTI_rolling_avg, aes(x = year)) +
  geom_line(aes(y = five_year_avg, color = "5 Year Avg")) +
  geom_line(aes(y = ten_year_avg, color = "10 Year Avg")) +
  geom_line(aes(y = twenty_year_avg, color = "20 Year Avg")) +
  labs(
    title = "Global Mean Temperature Anomaly Averages for 5, 10, and 20 Years",
    y = "Global Mean Temperature Anomaly",
    color = "Anomaly Averages")
```

The graph shows that each of the running averages has a different amplitude and makes that data seem to have more or less variance. The lower running averages have larger amplitudes while the higher running averages have lower amplitudes. This is because the higher running averages have more averages to draw from, thus making the average more accurate.

**Question 9:**

By examining the data here, what evidence do you find or not find for such a pause?

**Answer 9:**

From examining the data, I do not see any evidence of a pause since 1998, since the temperature anomaly seems to be increasing at a rate higher than ever before.

## Longer term trends in CO2 Records

When analyzing Earth’s climate, it is important to remember that Earth is 4.54 billion years old. Our analyses so far have only looked at recent history. How can we compare the recent data to prehistoric times? Are the current rates of change similar or different from those the earth has experienced in the past? To explore this, we can use data from ice cores drilled at the poles.

Hundreds of ice cores have been extracted from polar ice because they contain valuable data on atmospheric chemistry over pre-historic times. These valuable data exist in tiny air bubbles trapped in the ice. These air bubbles contain the same gases in the same ratios as the atmosphere at the time when the ice formed. The data you will analyze today are from ice cores extracted from the Vostok research station in Antarctica. As you have probably assumed, the depth of the ice core is related to how old the ice is; deep ice is older. There are two other variables that you will be analyzing from the ice cores. You will analyze CO2 concentration, which has been measured from air bubbles trapped in the ice. We can use these data to see what rates of change were like during this pre-historic period, during which human activity was minimal.

[Ice core data](https://data.ess-dive.lbl.gov/view/doi%3A10.3334%2FCDIAC%2FATG.009):

Vostok Core, back to 400,000 years before the present day

-   Description of data set: <https://data.ess-dive.lbl.gov/view/doi%3A10.3334%2FCDIAC%2FATG.009>
-   data: <https://data.ess-dive.lbl.gov/catalog/d1/mn/v2/object/ess-dive-457358fdc81d3a5-20180726T203952542>

You will use this code to download the data to your computer.

```{r}
download.file("https://data.ess-dive.lbl.gov/catalog/d1/mn/v2/object/ess-dive-457358fdc81d3a5-20180726T203952542",
              destfile = "vostok.icecore.co2")
```

You can click on the file in your Files pane to view it before reading into R.

**Question 10:**

The broad question is: how do recent atmospheric CO2 levels compare to historical levels?

Your answer to Question 10 is going to be a mix of code chunks and text that you put below in "Answer 10:"

-   Describe the data set: what are the columns and units? Where do the numbers come from?
-   What is the temporal resolution of the data?
-   Read in and prepare data for analysis.\
-   Reverse the ordering to create a chronological record so that each measurement is associcated with calender year.
-   Plot data.\
-   Combine this time series with the Mauna Loa data (the Warm-up exercise in this assignment). You will use "bind_rows()".
-   Plot the combined data. (the most recent time period must be on the right side of the plot).\
-   Describe your conclusions to the question "How do recent atmospheric CO2 levels compare to historical levels?" using your plot as supporting evidence.

**Answer 10:**

-   The data set contains data on ice cores recovered in Antarctica in 1998. The columns are Depth (of ice cores in m), age of the ice (year before present), mean age of the air (year before present), and CO2 concentration (ppmv).

-   ?

-   Read in and prepare data below

```{r}
icecore_co2 <-  read_table("vostok.icecore.co2", 
                  comment="#",
                  skip = 20,
                  col_names = c("depth", "ice_age", "mean_air_age", "co2_conc"))

icecore_co2
```

-   Reverse the order

    ```{r}
    icecore_co2 <- icecore_co2 |> 
      arrange(desc(ice_age))
    ```

-   Plot data

    ```{r}
    icecore_co2_plot <- ggplot(data = icecore_co2, aes(x = ice_age, y = co2_conc))+
      geom_line() +
      labs(
        title = "Ice Core Co2 over Time",
        x = "Ice Age (Year BP)",
        y = "Co2 Concentration (ppm)")

    icecore_co2_plot
    ```

-   Bind Mauna Loa Data

    ```{r}
    co2_combined <- bind_rows(icecore_co2, co2)
    ```

-   Plot

    ```{r}
    co2_combined_plot <- ggplot(data = co2_combined, aes(x = ice_age, y = co2_conc))+
      geom_line() +
      labs(
        title = "Ice Core Co2 over Time",
        x = "Ice Age (Year BP)",
        y = "Co2 Concentration (ppm)")

    icecore_co2_plot
    ```

-   "How do recent atmospheric CO2 levels compare to historical levels?" using your plot as supporting evidence.?

    **answer**

# Render and committing

Remember to Render your document as HTML and comment+push to GitHub your code and rendered HTML that was created when you knitted the document. Your GitHub repository should have multiple commits with informative commit messages.

# Attribution

Include citation of any AI-generated assistance or discussion with classmates (per policy in syllabus). Proper documentation of AI-generated assistance includes the prompt, the source (e.g., ChatGPT), and the significant parts of the response. Proper documentation of discussion with classmates include listing their names and the components discussed.

**Q7:**
